---
title: "**Comparing the AMM-side T-learner against External Meta-learners**"
subtitle: "Operational recipe: `grf` R-side and EconML Python-side adapters (Sub-phase 8.5.B)"
author: "**José Mauricio Gómez Julián**"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    theme: cosmo
    highlight: tango
    df_print: kable
    mathjax: default
  word_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Comparing the AMM-side T-learner against External Meta-learners}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

# **1. What this vignette covers**

This is an operational recipe. It walks you through `gdpar_compare_meta_learners()`
end to end with two reference adapters: `gdpar_adapter_grf()` (R-side,
based on the `grf` package) and `gdpar_adapter_econml()` (Python-side,
based on the `econml` library via `reticulate`). It also shows how to
extend the comparator by writing your own adapter, using `DoubleML` as
the worked example, and ends with troubleshooting for the
Python-side path.

The theoretical canonization of the comparator — definitions, the
adapter contract, the concordance criterion, identifiability per arm
under cross-method comparison, and the limits of the exercise — lives
in the companion canonical vignette
`vignette("v08c_meta_learner_comparison")`. Read that one if you want
to know *why* a given choice was made; read this one if you want to
*do* it.

We assume you have a `gdpar_causal_bridge` already built from a pair
of `gdpar_fit` objects (one per arm). If you are new to the bridge,
see `vignette("v08b_cate_ite_bridge_implementation")` first.

---

# **2. Setup**

Synthetic data, two arms, two fits, one bridge. The example is
deliberately small so the chunk runs in a few seconds.

```{r setup-data, eval = FALSE}
library(gdpar)
set.seed(2026L)
n <- 300L

df <- data.frame(x1 = rnorm(2L * n))
df$arm <- rep(c("treat", "ctrl"), each = n)
df$y <- with(df, ifelse(arm == "treat", 0.5, 0) +
                 0.8 * x1 +
                 rnorm(2L * n, sd = 0.5))
df_t <- subset(df, arm == "treat"); df_t$arm <- NULL
df_c <- subset(df, arm == "ctrl");  df_c$arm <- NULL

fit_t <- gdpar(y ~ x1, amm = amm_spec(a = ~ x1), data = df_t,
               iter_warmup = 300, iter_sampling = 300, chains = 2)
fit_c <- gdpar(y ~ x1, amm = amm_spec(a = ~ x1), data = df_c,
               iter_warmup = 300, iter_sampling = 300, chains = 2)

newdata <- data.frame(x1 = seq(-2, 2, length.out = 21L))
bridge <- gdpar_causal_bridge(fit_t, fit_c, newdata = newdata)
```

From here on we have a `bridge` of class `gdpar_causal_bridge`. The
comparator takes that object and a list of adapters and never touches
the two fits again.

---

# **3. The `grf` adapter in three lines**

`grf` is in `Suggests`. The adapter is constructed with
`gdpar_adapter_grf()`; you only need to pass the hyperparameters you
want to override. Sensible defaults match `grf`'s own defaults
(`num_trees = 2000L`, honesty on).

```{r grf-quick, eval = FALSE}
adapter_grf <- gdpar_adapter_grf(num_trees = 500L, seed = 2026L)
cmp <- gdpar_compare_meta_learners(
  bridge,
  methods = list(grf = adapter_grf)
)
print(cmp)
summary(cmp)
```

What you should see in `print(cmp)`:

- A line per method (only `grf` here) with `native_ci = TRUE`,
  the wall-clock time, and `predict = TRUE` (the adapter exposes
  `predict_fun`, so re-evaluating on a fresh grid does not refit).
- Three small concordance matrices indexed by `bridge` and `grf`:
  RMSE, Pearson, MAD.

If you want to re-evaluate the comparison on a fresh grid without
refitting `grf`:

```{r grf-predict, eval = FALSE}
newdata2 <- data.frame(x1 = seq(-1.5, 1.5, length.out = 15L))
cmp_new  <- predict(cmp, newdata = newdata2)
```

`predict.gdpar_meta_learner_comparison` reuses the cached `state`
inside `cmp$external$grf$state` (the fitted `grf::causal_forest`
object). The bridge is re-evaluated on `newdata2` via the embedded
`gdpar_causal_bridge` and the new concordance matrices are recomputed.

---

# **4. The EconML adapter**

`reticulate` and the Python module `econml` are both optional.
`reticulate` is in `Suggests`; `econml` is a Python package that lives
in your active Python environment. The package does not install Python
dependencies on your behalf.

## **4.1. One-time installation**

The recommended flow:

```{r econml-install, eval = FALSE}
# 1. Install reticulate (R-side) if absent.
install.packages("reticulate")

# 2. Register econml as a Python requirement, then install it.
reticulate::py_require("econml")  # reticulate 1.46+ ephemeral-env style
reticulate::py_install("econml")  # adds econml to the active env

# 3. Verify.
reticulate::py_module_available("econml")  # should be TRUE
```

The `py_require` call is a no-op on reticulate releases that predate
the ephemeral-env management; on 1.46 and later it tells reticulate
which Python module to pin in the active uv-managed environment.
`py_install` then performs the actual installation when needed.

If you maintain a virtualenv or conda env explicitly, install
`econml` in it with `pip install econml` or
`conda install -c conda-forge econml` and point `reticulate` to it
via `RETICULATE_PYTHON` or `reticulate::use_virtualenv()` / `use_condaenv()`.

## **4.2. Running the EconML adapter**

Once the Python module is available, the adapter is constructed
exactly like the `grf` one:

```{r econml-use, eval = FALSE}
adapter_econml <- gdpar_adapter_econml(n_estimators = 500L, seed = 2026L)
cmp2 <- gdpar_compare_meta_learners(
  bridge,
  methods = list(grf    = adapter_grf,
                 econml = adapter_econml)
)
print(cmp2)
summary(cmp2)
```

The concordance matrices now have three rows / columns (`bridge`,
`grf`, `econml`). Both adapters expose `predict_fun`, so
`predict(cmp2, newdata = newdata_fresh)` reuses both cached states
without refitting either model.

## **4.3. Caveat: serialization of the EconML state**

The `state` slot for the EconML adapter holds a reference to a Python
object managed by `reticulate`. The reference is valid for the
duration of the R session in which the comparison was built;
`saveRDS(cmp2, file = ...)` and a fresh-session `readRDS` will lose
that reference, and a subsequent call to `predict(cmp2_restored, ...)`
aborts cleanly with `gdpar_unsupported_feature_error`. Either re-fit
in the new session or build the comparison there from scratch. The
`grf` state survives serialization without modification.

---

# **5. Reading the output**

The print method emits four blocks:

```
<gdpar_meta_learner_comparison>
  n_obs                : <number of evaluation rows>
  n_methods (external) : <length of methods>
  level                : <inherited from bridge>
  methods              :
    - grf            native_ci = TRUE  time = ...  notes = 0  predict = TRUE
    - econml         native_ci = TRUE  time = ...  notes = 0  predict = TRUE

  concordance matrices (m-by-m, m = 1 + n_methods):
    RMSE: <symmetric m-by-m matrix>
    Pearson: <symmetric m-by-m matrix>
    MAD: <symmetric m-by-m matrix>
```

- A small RMSE between `bridge` and an external method indicates
  point-estimate agreement on `cate_mean` across the evaluation grid.
- A high Pearson with small RMSE means the two methods agree both in
  shape and in absolute scale. A high Pearson with large RMSE means
  the two methods agree on the shape of the CATE surface but disagree
  on the level (e.g. one is shifted).
- MAD is more robust than RMSE in the presence of a few outliers.

The summary method (`summary(cmp)`) returns a structured object with
three slots: `ate_table` (one row per method with the marginal ATE
and CI bounds), `metrics` (the long-format version of the three
matrices), `timing` (a per-method timing table). Use it for tables in
reports.

---

# **6. Writing your own adapter (DoubleML as an example)**

The contract of an adapter is two functions:

- `fit_predict_fun(X, Y, T, X_newdata, level, seed_run)` returns
  `list(cate_mean, cate_ci, state, notes)`.
- `predict_fun(state, X_newdata, level)` (optional) returns
  `list(cate_mean, cate_ci)`. When absent, `predict()` falls back to
  `fit_predict_fun` and emits a `gdpar_diagnostic_warning`.

A worked sketch with `DoubleML` (R-side; install with
`install.packages("DoubleML")`):

```{r doubleml-sketch, eval = FALSE}
fit_predict_dml <- function(X, Y, T, X_newdata, level, seed_run) {
  if (!requireNamespace("DoubleML", quietly = TRUE) ||
      !requireNamespace("mlr3learners", quietly = TRUE)) {
    stop("DoubleML and mlr3learners are required for this adapter.")
  }
  d <- cbind(X, Y = as.numeric(Y), T = as.integer(T))
  dml_data <- DoubleML::DoubleMLData$new(d, y_col = "Y", d_cols = "T",
                                          x_cols = setdiff(colnames(d),
                                                            c("Y", "T")))
  learner_g <- mlr3::lrn("regr.ranger", num.trees = 200L)
  learner_m <- mlr3::lrn("classif.ranger", num.trees = 200L,
                          predict_type = "prob")
  model <- DoubleML::DoubleMLPLR$new(dml_data, ml_g = learner_g$clone(),
                                      ml_m = learner_m$clone())
  model$fit()
  est <- as.numeric(model$coef)
  est_se <- as.numeric(model$se)
  z <- stats::qnorm(1 - (1 - level) / 2)
  n_new <- nrow(X_newdata)
  list(
    cate_mean = rep(est, n_new),
    cate_ci   = cbind(lower = rep(est - z * est_se, n_new),
                       upper = rep(est + z * est_se, n_new)),
    state     = list(model = model),
    notes     = "DoubleMLPLR returns a single ATE coefficient; broadcast to a constant CATE."
  )
}

predict_dml <- function(state, X_newdata, level) {
  n_new <- nrow(X_newdata)
  est <- as.numeric(state$model$coef)
  est_se <- as.numeric(state$model$se)
  z <- stats::qnorm(1 - (1 - level) / 2)
  list(
    cate_mean = rep(est, n_new),
    cate_ci   = cbind(lower = rep(est - z * est_se, n_new),
                       upper = rep(est + z * est_se, n_new))
  )
}

adapter_dml <- gdpar_meta_learner_adapter(
  name = "doubleml_plr",
  fit_predict_fun = fit_predict_dml,
  predict_fun = predict_dml,
  requires_r = c("DoubleML", "mlr3", "mlr3learners"),
  native_ci = TRUE,
  description = "DoubleMLPLR (constant CATE; useful as a robust ATE benchmark)"
)

cmp_with_dml <- gdpar_compare_meta_learners(
  bridge,
  methods = list(grf = adapter_grf, dml = adapter_dml)
)
```

A few notes on this sketch:

- `DoubleMLPLR` returns a *single* coefficient (an ATE), so the
  `cate_mean` is broadcast as a constant vector. That is honest: the
  partially linear model does not estimate a heterogeneous effect.
- If you want heterogeneity from `DoubleML`, use
  `DoubleMLIRM` with an interactive design and post-process the
  effect surface as you see fit; the contract is the same.
- The `notes` slot is the right place to document such quirks; the
  comparator surfaces it through `print()` and `summary()`.

---

# **7. Troubleshooting (Python-side)**

A short catalogue of what tends to go wrong and what to do.

**(a) `reticulate` not installed.** Install it with
`install.packages("reticulate")`. The package itself is small;
the heavy lifting is on the Python side.

**(b) `econml` not available.** Run
`reticulate::py_install("econml")`. On reticulate 1.46 and later,
reticulate uses ephemeral environments managed by `uv`; you may
need to call `reticulate::py_require("econml")` to pin the package
in the active environment before any Python operation.

**(c) Wrong Python detected.** Inspect `reticulate::py_config()` and
override with `Sys.setenv(RETICULATE_PYTHON = "/path/to/python")`
before loading reticulate, or use the explicit
`reticulate::use_virtualenv(...)`. If you maintain a Conda
environment, `reticulate::use_condaenv("name")` does the same.

**(d) `numpy` complains about version conflicts.** Pin the numpy
version your `econml` was built against:
`reticulate::py_install("numpy==1.26.*", pip = TRUE)`. EconML 0.16
is known to work with numpy 1.26.x as of mid-2026.

**(e) Cached Python state lost after restart.** The Python objects
inside `cmp$external$econml$state` do not survive R session
restarts. Rebuild the comparison in the new session.

**(f) Adapter aborts with `gdpar_missing_dependency_error`.** That
error is the package's deliberate, structured signal that a
`Suggests` package or a Python module is missing. The error
message names the missing item; install it and rerun. The
package never installs anything on your behalf.

---

# **8. Where to go next**

- `vignette("v08c_meta_learner_comparison")` — the canonical
  theoretical addendum (definitions, identification under
  cross-method comparison, the concordance criterion, the limits of
  the exercise, identifiability per arm under the bridge).
- `vignette("v08b_cate_ite_bridge_implementation")` — the canonical
  T-learner AMM-side bridge (the object you feed into the
  comparator).
- `vignette("v08_cate_ite_positioning")` — the positioning of the
  package's CATE / ITE workflow within the meta-learner literature.

---

*End of Operational Vignette -- Sub-phase 8.5.B.*
