---
title: "**The Empirical-Bayes Workflow in gdpar**"
subtitle: "Operational recipe: `gdpar_eb()` end-to-end, four path regimes, EB-vs-FB comparison and troubleshooting (Sub-phase 8.6.E)"
author: "**José Mauricio Gómez Julián**"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    theme: cosmo
    highlight: tango
    df_print: kable
    mathjax: default
  word_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The Empirical-Bayes Workflow in gdpar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>",
  eval = FALSE
)
```

# **1. What this vignette covers**

This is an operational recipe. It walks you through `gdpar_eb()` end
to end across the four path regimes canonized in Sub-phase 8.6 of
gdpar:

| Regime  | $K$ | $p$ | Stan template pair                                        |
|:--------|:---:|:---:|:----------------------------------------------------------|
| Base    |  1  |  1  | `amm_eb_marginal.stan` + `amm_eb_conditional.stan`        |
| Path A  |  1  | >1  | `amm_eb_marginal_multi.stan` + `amm_eb_conditional_multi.stan` |
| Path B  | >1  |  1  | `amm_eb_marginal_K.stan` + `amm_eb_conditional_K.stan`    |
| Path C  | >1  | >1  | `amm_eb_marginal_KxP.stan` + `amm_eb_conditional_KxP.stan`|

It then shows how to compare an EB fit against a Fully-Bayes (FB) fit
on the same data via `gdpar_compare_eb_fb()`, how to read the
numerical diagnostics that surface when the marginal Hessian is
poorly conditioned, and when to give up on EB and fall back to FB.

The theoretical canonization of EB-vs-FB — the asymptotic
equivalence (Theorem 7A and its multivariate extension Theorem 7A\*),
the higher-order coverage discrepancy (Proposition 7B scalar /
7B\* matricial / 7B\* tensorial), the finite-sample compound decision
bound (Theorem 7C / 7C\* compound multi-slot), and the four
discrepancy conditions of Proposition 7D — lives in the canonical
vignettes `vignette("v07_eb_vs_fb")` (scalar) and
`vignette("v07b_eb_multivariate")` (multivariate extension). Read
those if you want to know *why* a given choice was made; read this
one if you want to *do* it.

The chunks below default to `eval = FALSE` because they compile Stan
models and take several minutes per fit; re-enable evaluation on a
per-chunk basis or globally via `knitr::opts_chunk$set(eval = TRUE)`
if you want to reproduce the runs locally.

# **2. Setup**

```{r packages}
library(gdpar)
# These two are Suggests; gdpar_eb() and gdpar_compare_eb_fb() require
# them at runtime.
library(cmdstanr)
library(posterior)
```

We will work with a synthetic dataset of size `n = 150` on a single
continuous outcome with one covariate, then enrich it to multivariate
and multi-slot variants as we walk through the four regimes.

```{r data}
set.seed(20260526L)
n  <- 150L
df <- data.frame(x = stats::rnorm(n))
df$y_scalar <- 1.0 + 0.4 * df$x + stats::rnorm(n, sd = 0.3)
# Multivariate (p = 2) outcome for Path A.
df$y_p2 <- cbind(
  1.0 + 0.4 * df$x + stats::rnorm(n, sd = 0.3),
  -0.5 + 0.2 * df$x + stats::rnorm(n, sd = 0.4)
)
# Same dataset is fine for Path B (K > 1, p = 1) and Path C (K > 1,
# p > 1) by reusing y_scalar / y_p2 with a multi-slot family below.
```

# **3. Minimal `gdpar_eb()` call (K = 1 + p = 1)**

The base regime mirrors the canonical `gdpar()` signature; the only
new arguments are `eb_correction = TRUE` (default; applies the
Proposition 7B scalar inflation to the conditional credible intervals)
and `laplace_control = list(...)` (controls the multi-start Laplace
maximizer of v07 Section 11.1 step (i)).

```{r eb-base}
fit_eb <- gdpar_eb(
  formula        = y_scalar ~ x,
  family         = gdpar_family("gaussian"),
  amm            = amm_spec(a = ~ x),
  data           = df,
  iter_warmup    = 500L,
  iter_sampling  = 500L,
  chains         = 2L,
  refresh        = 0L,
  seed           = 1L,
  laplace_control = list(multi_start_M = 5L)
)
print(fit_eb)
```

The output reports the EB plug-in point estimate
$\widehat\theta_{\text{ref}}^{\text{EB}}$ (from the marginal Laplace),
its marginal standard error from the Laplace covariance, the marginal
Hessian condition number $\kappa(H)$, the multi-start dispersion
across the `multi_start_M = 5` independent inits, and the conditional
HMC convergence diagnostics ($\widehat R$, ESS, divergences).

`summary(fit_eb)` returns a tidy table of EB credible intervals with
the Proposition 7B scalar inflation applied:

```{r eb-base-summary}
summary(fit_eb)
```

# **4. Path A (K = 1, p > 1): multivariate outcome**

For a $p$-dimensional outcome, the amm spec uses `dimwise()` (or a
plain `list` of length $p$) to declare the per-coordinate components,
and the family is promoted to a `gdpar_family_multi` of dimension
$p$. The Stan template pair `amm_eb_marginal_multi.stan` +
`amm_eb_conditional_multi.stan` (canonized in Sub-phase 8.6.C under
decision D34) is dispatched automatically:

```{r eb-path-A}
fit_eb_A <- gdpar_eb(
  formula        = y_p2 ~ x,
  family         = gdpar_family_multi("gaussian", p = 2L),
  amm            = amm_spec(p = 2L, dims = dimwise(a = ~ x)),
  data           = df,
  iter_warmup    = 500L,
  iter_sampling  = 500L,
  chains         = 2L,
  refresh        = 0L,
  seed           = 2L
)
print(fit_eb_A)
```

The corresponding correction is matricial, $C^*_{g,\alpha} \in
\mathbb{R}^{p\times p}$ (Proposition 7B\* of v07b Section 5.1). It
reduces algebraically to the scalar Proposition 7B at $p = 1$, so the
upgrade from base to Path A is transparent.

# **5. Path B (K > 1, p = 1): multi-parametric family**

For a multi-slot distributional regression (e.g. modelling both `mu`
and `sigma` of a Gaussian K=2, or `mu` and `phi` of a Negative
Binomial K=2), the amm input is either a named list of `amm_spec`
(one per slot) or a `gdpar_formula_set` via `gdpar_bf(...)`. The
Stan template pair `amm_eb_marginal_K.stan` +
`amm_eb_conditional_K.stan` is dispatched automatically:

```{r eb-path-B}
fs <- gdpar_bf(y_scalar ~ a(x), sigma ~ a(x))
fit_eb_B <- gdpar_eb(
  formula        = fs,
  family         = gdpar_family("gaussian"),
  data           = df,
  iter_warmup    = 500L,
  iter_sampling  = 500L,
  chains         = 2L,
  refresh        = 0L,
  seed           = 3L,
  skip_id_check  = TRUE
)
print(fit_eb_B)
```

Coverage of stan_ids per Sub-phase 8.6.C decision D33 (relaxed):
`{1, 3, 5, 6, 7, 8, 9, 10, 11, 12, 13}` — Gaussian, Negative
Binomial, Beta, Gamma, Lognormal-loc-scale, Student-t, Tweedie, ZIP,
ZINB, Hurdle-Poisson, Hurdle-NB.

# **6. Path C (K > 1, p > 1): full K x p extension**

Path C, canonized in Sub-phase 8.6.D under decision D36 = (alpha) +
D37 = (i) + D38'' = (h), composes Path A coordinate-wise
factorization with Path B multi-parametric K-slot semantics: a single
outcome matrix-column `y[n, p]` is shared across the K distributional
slots, each carrying its own per-coordinate linear predictor.

The initial 8.6.D iteration restricts Path C to
`family$stan_id %in% c(1, 3)` (Gaussian K=2, NB K=2) under decision
D40' to avoid the numerical caveat of Section 6.1 of the opening
handoff (HMC condicional bajo plug-in EB cerca del borde de soporte
logit/log links + warmup corto). Beta / Gamma / Lognormal /
Student-t / Tweedie / mixtures are deferred to a later iteration.

```{r eb-path-C}
y_p2_int <- matrix(
  rnbinom(n * 2L, size = 5, mu = exp(0.5 + 0.2 * df$x)),
  n, 2L
)
df$y_p2_int <- y_p2_int
fit_eb_C <- gdpar_eb(
  formula        = y_p2_int ~ x,
  family         = gdpar_family("neg_binomial_2"),
  amm            = list(
    mu  = amm_spec(p = 2L, dims = dimwise(a = ~ x)),
    phi = amm_spec(p = 2L, dims = dimwise(a = ~ x))
  ),
  data           = df,
  iter_warmup    = 500L,
  iter_sampling  = 500L,
  chains         = 2L,
  refresh        = 0L,
  seed           = 4L,
  skip_id_check  = TRUE
)
print(fit_eb_C)
```

The fit object carries new Path C-specific slots:

- `theta_ref_kp_hat`: 3D numeric array of shape `[J_groups, K, p]`.
- `theta_ref_kp_se`: same shape; per-coordinate marginal standard
  errors derived from each slot's Laplace covariance block.
- `theta_ref_kp_cov_per_slot`: named list of `K` matrices each
  `p × p`; the per-slot blocks of the joint Laplace covariance after
  the block-diagonal extraction of decision D43 = (a).
- `correction_tensor_constant`: 3D array of shape `[K, p, p]` with
  the Proposition 7B\* tensor correction of decision D37 = (i).
- `correction_tensor_dispositions`: named character vector reporting
  whether each slot's correction was applied (`"ok"`),
  `"non_finite"`, `"non_psd"`, `"missing"`, or `"disabled"`.

# **7. Numerical diagnostics: how to read them**

Every `gdpar_eb_fit` carries a `diagnostics_numerical` slot with
four entries derived from the multi-start Laplace strategy of
Charter Section 2.8:

- `kappa` (base regime / Path A / Path B) or `kappa_per_slot` (Path C):
  the condition number of the marginal Hessian at the chosen MAP. The
  guard `laplace_control$kappa_threshold` (default `1e10`) aborts the
  fit with `gdpar_eb_numerical_error` when exceeded. A `kappa` between
  `1e6` and `1e9` is a warning sign; consider tightening the prior
  on `theta_ref` or moving the anchor closer to the data via
  `anchor = "empirical_y"`.

- `lm_perturbation`: the Levenberg-Marquardt ridge actually added to
  the marginal covariance when the bare Hessian was singular or
  non-PSD. A non-zero value means the Laplace approximation needed
  numerical stabilization; the result is still valid but the
  effective sample size for the EB anchor is smaller than the nominal
  Laplace draws would suggest.

- `multi_start_dispersion`: standard deviation of the log marginal
  across the `multi_start_M` independent inits, normalized by the
  absolute mean. A value above `0.05` triggers a diagnostic warning
  because it suggests multi-modality of the marginal likelihood (open
  question O5\*-EBFB of v07b Section 9.5). Consider raising
  `multi_start_M` from the default 5 to 10–20 to stress-test the
  optimum.

- `marginal_log_lik_history`: the per-init log marginal achieved by
  each `optimize()` call. Useful for forensics when the dispersion is
  large.

# **8. EB vs FB comparison via `gdpar_compare_eb_fb()`**

The companion function `gdpar_compare_eb_fb()` (canonized in
Sub-phase 8.6.E) takes a `gdpar_eb_fit` and a `gdpar_fit` fitted on
the same dataset and reports three operational diagnostics of the
EB-vs-FB theory of v07:

```{r compare-eb-fb}
fit_fb <- gdpar(
  formula        = y_scalar ~ x,
  family         = gdpar_family("gaussian"),
  amm            = amm_spec(a = ~ x),
  data           = df,
  iter_warmup    = 500L,
  iter_sampling  = 500L,
  chains         = 2L,
  refresh        = 0L,
  seed           = 1L
)
cmp <- gdpar_compare_eb_fb(fit_eb, fit_fb, level = 0.95,
                            tv_bins = 30L)
print(cmp)
summary(cmp)
```

The output carries three tables:

- `theta_diff_table`: per-anchor cell comparison of
  $\widehat\theta_{\text{ref}}^{\text{EB}}$ and the FB posterior mean
  $E_{\text{FB}}[\theta_{\text{ref}}]$, with the difference and the
  difference normalized by the FB standard error. Under the standing
  hypotheses of v07 Section 4 (EB-MARG-ID + PRIOR-FB-WEAK +
  HIER-COMPLEX), Theorem 7A predicts `diff_rel` close to zero up to
  $O(n^{-1/2})$.

- `tv_table`: marginal empirical total variation distance per
  common $\xi$ parameter, computed via histogram plug-in over the
  shared support. Under Theorem 7A, marginal TV $\to 0$ in
  probability as $n \to \infty$. A persistent large TV across many
  parameters suggests one of the discrepancy conditions of
  Proposition 7D (multi-modality of the marginal likelihood,
  near-singular Fisher information, informative prior, deep
  hierarchy).

- `coverage_table`: per anchor cell, the EB credible interval width
  (with inflation applied when `eb_correction = TRUE`), the FB
  credible interval width, and the ratio `width_eb / width_fb`. This
  operationally verifies the $O(n^{-1})$ under-cover prediction of
  Proposition 7B (and its matricial / tensorial extensions). A ratio
  systematically below 1 confirms the EB under-cover; a ratio above
  1 after correction suggests the inflation is over-correcting,
  which is acceptable in finite samples and consistent with the
  asymptotic guarantee.

# **9. Troubleshooting**

## 9.1. `gdpar_eb_numerical_error: kappa = ...`

The marginal Hessian is too ill-conditioned for Laplace to be
reliable. Options:

1. Tighten the prior on `theta_ref` via a stronger
   `gdpar_prior(theta_ref = "normal(0, 0.5)")`.
2. Move the anchor closer to the data via
   `anchor = "empirical_y"`.
3. Raise `laplace_control$multi_start_M` and rerun (the per-init
   seed offset will sample different unconstrained-space inits).
4. Fall back to FB via `gdpar()` (the most reliable option when the
   marginal likelihood is genuinely flat).

## 9.2. `gdpar_unsupported_feature_error` on Path C

The initial 8.6.D iteration restricts Path C to
`family$stan_id %in% c(1, 3)` (Gaussian K=2, NB K=2) per decision
D40'. For other distributional families under $K > 1 \wedge p > 1$,
fall back to FB via `gdpar()` or split the multivariate outcome into
$p$ separate Path B fits as a workaround.

## 9.3. Multi-start dispersion warning

A `multi_start_dispersion` above 0.05 suggests multi-modality of the
marginal likelihood (open question O5\*-EBFB of v07b Section 9.5).
The result returned by `gdpar_eb()` is the mode with the highest log
marginal across the `multi_start_M` inits, which may differ from the
mode picked by a single run. Options:

1. Raise `multi_start_M` from the default 5 to 10–20.
2. Use a stronger prior on `theta_ref` to break the multi-modality.
3. Fall back to FB; HMC explores multi-modal posteriors better than
   single-mode Laplace.

## 9.4. Path B conditional HMC instability under `logit`-strict links

Path B with strict $(0, 1)$ inverse-link families (Beta, Bernoulli)
can exhibit conditional HMC instability when the EB plug-in anchor
falls near the support boundary and the HMC warmup is short. The
caveat is documented in the closure of Sub-phase 8.6.C
(`HANDOFF_SUBFASE_8_6_C_CIERRE.md` Section 3.3). Workarounds:

1. Raise `iter_warmup` to 1000 or more.
2. Use a tighter prior on `sigma_a_k` (the HMC has fewer chances to
   sample large `a_coef_k` values that saturate the logit).
3. Use `anchor = "prior_median"` or an explicit numeric anchor that
   keeps the EB plug-in in the interior of the support.

## 9.5. Path C smoke validation

A representative end-to-end Path C smoke (Gaussian K=2 + p=2,
$n = 80$, `iter_warmup = iter_sampling = 200L`, `chains = 2L`)
takes approximately 50 seconds on contemporary hardware
(`amm_eb_marginal_KxP.stan` and `amm_eb_conditional_KxP.stan` each
compile once and are cached by cmdstanr). If the smoke takes
significantly longer or aborts with `gdpar_eb_numerical_error`, the
geometry of the K × p marginal is likely the culprit; check
`kappa_per_slot` and the slot dispositions returned by the
correction tensor.

# **10. Where to go next**

- For the theoretical canonization of EB-vs-FB at the scalar level,
  read `vignette("v07_eb_vs_fb")`.
- For the multivariate / multi-slot extension of the theory (Theorem
  7A\*, Proposition 7B\*, Theorem 7C\* compound multi-slot, open
  question O5\*-EBFB), read `vignette("v07b_eb_multivariate")`.
- For the AMM canonization on which both EB and FB operate, read
  the AMM family of canonical vignettes (v01–v06).
- For CATE / ITE applications and the EB-vs-FB shrinkage discussion
  in the context of heterogeneous treatment effects, read
  `vignette("v08_amm_for_cate_ite_positioning")` and the
  implementation vignettes `v08b_cate_ite_bridge_implementation`
  and `v08c_meta_learner_comparison`.
- If your data are dependent — **temporally** (a time series) or
  **spatially** — the EB posterior understates uncertainty because the
  likelihood assumes conditional independence. Measure the violation
  with `gdpar_dependence_diagnostic()` (lag-1 autocorrelation,
  Durbin-Watson, Ljung-Box) or its spatial sibling
  `gdpar_spatial_dependence_diagnostic()` (Moran's I), and obtain
  dependence-robust standard errors and percentile intervals with
  `gdpar_dependence_robust()` (temporal block bootstrap) or
  `gdpar_spatial_dependence_robust()` (spatial block bootstrap), Block
  9, Axis 2. This makes the *uncertainty* robust to dependence; gdpar
  does not *model* the dependence. The full recipe is in
  `vignette("vop09_dependence_robust")`.

# **References**

The canonical theoretical references for EB-vs-FB in the AMM are
Petrone, Rousseau, and Scricciolo (2014) and Rousseau and Szabo
(2017) for the asymptotic equivalence and merging results; Carlin
and Gelfand (1990) for the higher-order coverage discrepancy and the
inflation correction; and Robbins (1956) / Efron (2010) for the
compound decision framework. Full bibliographic details are in
`vignette("v07_eb_vs_fb")` Section "References Cited in This Block".
