---
title: "**Dependence-Robust Inference in gdpar (Axis 2)**"
subtitle: "Making the uncertainty honest under temporal and spatial dependence, without modelling it (Block 9, Axis 2)"
author: "**José Mauricio Gómez Julián**"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    theme: cosmo
    highlight: tango
    df_print: kable
    mathjax: default
  word_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Dependence-Robust Inference in gdpar (Axis 2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>",
  eval = FALSE
)
```

# **1. What this vignette covers**

gdpar's Path 1 estimates — Empirical-Bayes (`gdpar_eb`) or full-Bayes
(`gdpar`) — assume the observations are **conditionally independent** given the
mean structure. When the data carry **temporal** (serial) or **spatial**
autocorrelation, that assumption is violated: the point estimates remain
consistent if the mean structure is correctly specified, but the model-based
(posterior / Laplace) standard errors are **too narrow** and the intervals
**under-cover**. Sections 2–3 use the Empirical-Bayes path; Section 4 shows the
identical workflow on a full-Bayes fit.

Axis 2 of Block 9 is the *inferential* response to this. It is deliberately
**not** a model of the dependence:

> **Honest scope.** gdpar does **not** model the correlated noise or the spatial
> random field (that is Axis 1 / a later modelling block, deferred and
> evidence-gated). Axis 2 only makes the *inference* robust to dependence. The
> stance is the working-independence + robust-variance estimator of
> Liang & Zeger (1986): the **point estimates are unchanged**; only the reported
> uncertainty is re-estimated to be dependence-robust.

Two symmetric pairs of functions, sharing one internal refit engine:

| Dependence | Diagnostic | Robust SE / intervals |
|---|---|---|
| **Temporal** | `gdpar_dependence_diagnostic()` | `gdpar_dependence_robust()` |
| **Spatial** | `gdpar_spatial_dependence_diagnostic()` | `gdpar_spatial_dependence_robust()` |

The workflow is always the same: **diagnose first**, and only if dependence is
flagged, **re-estimate the uncertainty**.

# **2. The temporal pair (recap)**

```{r temporal}
library(gdpar)

set.seed(1)
n <- 150
x <- rnorm(n)
# AR(1) errors: serial dependence the working-independence fit ignores.
y <- 1 + 0.5 * x + as.numeric(stats::arima.sim(list(ar = 0.6), n))
df <- data.frame(x = x, y = y, t = seq_len(n))

fit <- gdpar_eb(y ~ x, amm = amm_spec(a = ~ x), data = df,
                chains = 2, iter_warmup = 300, iter_sampling = 300)

# Step 1 — diagnose. Lag-1 autocorrelation, Durbin-Watson, Ljung-Box.
gdpar_dependence_diagnostic(fit, index = df$t)

# Step 2 — if flagged, re-estimate the uncertainty by a temporal block bootstrap.
gdpar_dependence_robust(fit, data = df, index = df$t, B = 199, seed = 1)
```

The robust table reports, per coefficient, the `estimate` (unchanged), the
`model_se`, the bootstrap `robust_se`, their ratio `se_ratio` and percentile
`ci_lower` / `ci_upper`. A `se_ratio > 1` means the model-based SE understates
the dependence-robust SE.

**The block length.** By default `block_length = NULL` uses the rate-optimal
`round(n^(1/3))` (it fixes only the *rate*). You may fix an integer manually, or
let the data choose it:

```{r temporal-auto}
# Data-driven block length: the Politis-White (2004) automatic selector,
# computed from the residuals (no extra refit), with the rate as fallback.
gdpar_dependence_robust(fit, data = df, index = df$t,
                        block_length = "auto", B = 199, seed = 1)
```

`"auto"` runs the canonical Politis & White (2004) rule (Patton, Politis & White
2009 correction): it reads the residual autocorrelations, finds the lag beyond
which they are negligible, and returns
`b_opt = (2 ĝ² / D)^(1/3) n^(1/3)` with the overlapping-block constant
`D = (4/3) spec²`. Stronger serial dependence gives a longer block; white-noise
residuals give unit blocks; a degenerate series falls back to the rate. The
chosen value and method are reported in `block_length` and
`block_length_method`. (The selector assumes the fitted parameter count is small
relative to `n`.)

# **3. The spatial pair**

The spatial functions replace exactly two pieces of the temporal machinery — the
**statistic** (lag-1 autocorrelation becomes **Moran's I**) and the
**resampling** (1-D contiguous blocks become **2-D spatial blocks**) — and reuse
everything else.

## 3.1. Diagnostic: Moran's I

```{r spatial-diagnostic}
set.seed(2)
n <- 200
gx <- runif(n); gy <- runif(n)            # spatial coordinates
x <- rnorm(n)
# An omitted smooth spatial trend lands in the residuals.
y <- 1 + 0.5 * x + 3 * (gx + gy) + rnorm(n, sd = 0.3)
df <- data.frame(x = x, y = y)

fit_sp <- gdpar_eb(y ~ x, amm = amm_spec(a = ~ x), data = df,
                   chains = 2, iter_warmup = 300, iter_sampling = 300)

gdpar_spatial_dependence_diagnostic(fit_sp, coords = cbind(gx, gy), seed = 1)
```

Moran's I is

$$
I \;=\; \frac{n}{S_0}\,
        \frac{\sum_i \sum_j w_{ij}\,(r_i - \bar r)(r_j - \bar r)}
             {\sum_i (r_i - \bar r)^2},
\qquad S_0 = \sum_i \sum_j w_{ij},
$$

with $r$ the residuals and $W = (w_{ij})$ a **spatial weight matrix**. Under the
null of spatial exchangeability $\mathbb{E}[I] = -1/(n-1)$.

**Choosing the weights.** By default gdpar builds a **row-standardized
$k$-nearest-neighbour** graph (so $S_0 = n$), with the declared heuristic
$k = \max(4, \min(\lfloor \log n \rceil, n-1))$ — robust to irregular spacing and
guaranteeing no isolated point. Alternatives:

* `weights = "distance"` — a distance band whose threshold is the smallest that
  isolates no location.
* `W = <your matrix>` — supply your own neighbourhood (adjacency, flow network,
  administrative contiguity). This is the right choice whenever domain knowledge
  defines the structure; it is row-standardized internally.

**Significance.** The default is a **two-sided permutation test** (`n_perm = 999`):
the residuals are relabelled across locations, $I$ is recomputed, and
$p = (1 + \#\{|I_{\text{perm}} - \mathbb{E}[I]| \ge |I - \mathbb{E}[I]|\}) /
(n_{\text{perm}} + 1)$. It assumes neither normal residuals nor a symmetric $W$,
so it is safe with Dunn-Smyth residuals and the asymmetric kNN graph. The
analytic Cliff-Ord normal approximation is available via `test = "analytic"`
(cheaper, but it warns under an asymmetric $W$).

## 3.2. Robust SE: spatial block bootstrap

```{r spatial-robust}
gdpar_spatial_dependence_robust(fit_sp, data = df, coords = cbind(gx, gy),
                                B = 199, seed = 1)
```

The bounding box of `coords` is tiled into a `g × g` grid; non-empty cells are
resampled with replacement and concatenated to length `n`. By default the grid
**origin is randomized** per replicate (`random_origin = TRUE`,
Politis-Romano-Lahiri), which breaks the deterministic cell-boundary artifact.
An overlapping `scheme = "moving"` is also available. The output table has the
same columns as the temporal one.

## 3.3. The default block size, and why `n^(1/4)`

The block side per axis defaults to `g = max(2, round(n^(1/4)))`. This is the
$d = 2$ case of the rate that minimises the **mean-squared error of the
block-bootstrap variance estimator**. Writing $M$ for the points per block (so
the block has linear extent $M^{1/d}$ per axis):

* bias from dependence broken at block edges: $O(M^{-1/d})$ (Künsch 1989; Hall,
  Horowitz & Jing 1995),
* estimator variance: $O(M/n)$,

so $\mathrm{MSE}(M) \sim M^{-2/d} + M/n$ is minimised at
$M \sim n^{\,d/(d+2)}$. At $d = 1$ this gives $M \sim n^{1/3}$ points per
block — **exactly** the temporal `block_length = round(n^(1/3))` default — and at
$d = 2$ it gives $M \sim n^{1/2}$, i.e. $g^2 = n/M \sim n^{1/2}$ cells, hence
$g \sim n^{1/4}$ cells per axis. The exponent is therefore the variance-optimal
rate that reduces correctly to the canonical temporal rate.

> **Registered dissent (D100).** A decorrelating cross-lineage review argued for
> the $n^{1/(d+4)}$ rate ($n^{1/6}$ at $d = 2$). That rate governs a *different*
> estimand — the second-order bias / two-sided distribution-function coverage,
> which gives $n^{1/5}$ at $d = 1$ and so does **not** reduce to the variance
> default's $n^{1/3}$. It is recorded as a dissent, not adopted.

## 3.4. Data-driven block size (`block_size = "auto"`)

The default above fixes only the *rate*. Its *constant* can be chosen from the
data — but, unlike the temporal case, Politis & White (2004) has **no**
established spatial plug-in (its flat-top spectral-density-at-zero estimator does
not extend cleanly to a field in the plane). gdpar therefore calibrates the
cells-per-axis `g` over a grid of candidates:

```{r spatial-auto}
gdpar_spatial_dependence_robust(fit_sp, data = df, coords = cbind(gx, gy),
                                block_size = "auto", B = 199, seed = 1)
```

For each candidate `g`, cheap (no-refit) spatial block resamples give the
bootstrap variance $V(g)$ of the **design-weighted residual functionals**
$(1/n)\,[1, \tilde{gx}, \tilde{gy}]^\top z$ — these are the influence directions
of the coefficient, so their optimal block size matches the coefficient's, which
the residual mean alone would not. The block size minimises an empirical
mean-squared error,

$$
g^\ast \;=\; \arg\min_g\;\Big[\underbrace{(\tilde V(g) - \tilde V(g_{\min}))^2}
                                       _{\text{squared bias}}
                            \;+\; \underbrace{c\,V(g)^2 / n_{\text{tiles}}(g)}
                                       _{\text{variance}}\Big],
$$

with the bias **anchored at the largest blocks** $g_{\min}$ (the least biased,
since the dependence-breaking bias of the variance estimator grows like
$g/\sqrt n$) and the variance scaling like the inverse number of blocks
(Lahiri 2003). Stronger / longer-range spatial dependence is captured by larger
blocks (smaller `g`); a short-range or near-independent field by smaller blocks
(larger `g`); a degenerate calibration falls back to the $n^{1/4}$ rate. The
choice and method are reported in `block_size` and `block_size_method`.

> **Provenance and honest scope (D101).** A decorrelating cross-lineage review
> supplied the empirical-MSE skeleton; two of its concrete choices — the bias
> anchor and the variance term — were **corrected after an audit and empirical
> validation**, because the proposed forms would have made the selector
> anticonservative (anchoring at the smallest blocks) or non-adaptive (a
> Monte-Carlo jackknife variance that vanishes with the resample count). A single
> **isotropic** `g` is used; strongly anisotropic residual dependence is a
> documented limitation (the minimal fix, two coordinate-wise calibrations, is
> deferred). As everywhere in Axis 2, this only sizes the resampling blocks — it
> does not model the dependence.

# **4. The full-Bayes path**

Everything above was shown on the Empirical-Bayes path (`gdpar_eb`). The exact
same four functions also accept a **scalar full-Bayes fit** (`gdpar`), so the
EB/FB asymmetry is closed: diagnose, then — if flagged — re-estimate, with no
change of API.

```{r fb, eval = FALSE, purl = FALSE}
# A full-Bayes fit instead of an Empirical-Bayes one:
fit_fb <- gdpar(y ~ x, amm = amm_spec(a = ~ x), data = df,
                chains = 2, iter_warmup = 300, iter_sampling = 300)

# Same diagnostic, same robust SE -- only the object class changes.
gdpar_dependence_diagnostic(fit_fb, index = df$t)
gdpar_dependence_robust(fit_fb, data = df, index = df$t, B = 199, seed = 1)
```

What differs is **internal**, and worth understanding before you read the table:

* **Point estimate and model SE.** On the full-Bayes path the per-coefficient
  point estimate is the **posterior mean** and `model_se` is the **posterior
  standard deviation** (vs. the Laplace mode / conditional posterior SD on the EB
  path). `robust_se` is, as always, the block-bootstrap SD of the per-refit point
  estimate, so `se_ratio = robust_se / model_se` stays a like-for-like SD-vs-SD
  ratio. The posterior mean / SD are chosen (over median / IQR) to keep that
  parity and to avoid an undeclared normal-scaling constant.
* **Cost.** Each refit re-runs the **full** HMC, so a full-Bayes
  `gdpar_dependence_robust()` is markedly more expensive than the Empirical-Bayes
  one. Keep `B` modest, and raise `iter_sampling` rather than `B` if the refit
  ESS (reported in the result's `refit_diagnostics`) is low — a finite-iteration
  refit slightly and *conservatively* inflates `robust_se`.
* **Reading `se_ratio` under an informative prior.** Unlike the EB path, a
  full-Bayes `se_ratio < 1` is usually **benign**: an informative prior
  concentrates the posterior beyond what the data alone support, so the posterior
  SD is smaller than the bootstrap SD even under correct independent
  specification (the conjugate-Gaussian algebra gives
  $\text{se\_ratio}^2 = n\tau/(n\tau+\tau_0) < 1$ with prior precision
  $\tau_0$). Only `se_ratio` clearly **above** 1 flags dependence /
  misspecification.
* **EB vs FB are different estimands for `theta_ref`.** The EB point estimate is
  the Laplace *mode*; the full-Bayes one is the posterior *mean*. They coincide
  asymptotically (Bernstein–von Mises) but may differ in finite samples, so the
  two paths' tables need not agree to the last digit on the same data.
* **Convergence accounting.** The result carries `refit_diagnostics` (aggregate
  `max_rhat`, `min_ess_bulk`, divergent / high-R-hat refit counts). Under-converged
  refits are **never excluded or down-weighted** — that screen would be
  non-random and would bias the SE — but a single note is emitted when a refit's
  R-hat clearly exceeds 1.05.

A *bagged* / widened posterior (BayesBag) would be a different object — a
re-architected estimator, not a robust variance for the same one — and is a
deliberately deferred lateral; Axis 2 keeps its honest scope, *robust variance,
not better estimates*, on both paths.

# **5. Reading the result, and the caveats**

* **`se_ratio > 1`** ⇒ the model-based uncertainty understated the
  dependence-robust uncertainty; trust the robust interval. **`se_ratio ≈ 1`** ⇒
  little correction needed.
* **Coordinates are Euclidean.** Project lon/lat first (e.g. to UTM); raw lon/lat
  distorts the neighbour graph, severely at high latitudes. Great-circle
  distance is deliberately not supported, to avoid a heavy **sf** / **geosphere**
  dependency.
* **Moran's I tests residual exchangeability, not its cause.** A significant $I$
  may reflect true spatial dependence **or** model misspecification (e.g. an
  omitted nonlinear effect). Interpret accordingly — a misspecified mean
  structure is better fixed than bootstrapped around.
* **Honest scope, restated.** The bootstrap delivers robust *variance*, not
  better point estimates, and is valid for weak / short-range dependence
  relative to the block size; it does not rescue long-memory or strong
  long-range dependence.

# **References**

* Cliff, A. D. & Ord, J. K. (1981). *Spatial Processes: Models and Applications*.
  Pion, London.
* Hall, P., Horowitz, J. L. & Jing, B.-Y. (1995). On blocking rules for the
  bootstrap with dependent data. *Biometrika* 82(3), 561-574.
* Künsch, H. R. (1989). The jackknife and the bootstrap for general stationary
  observations. *Annals of Statistics* 17(3), 1217-1241.
* Lahiri, S. N. (2003). *Resampling Methods for Dependent Data*. Springer.
* Liang, K.-Y. & Zeger, S. L. (1986). Longitudinal data analysis using
  generalized linear models. *Biometrika* 73(1), 13-22.
* Patton, A., Politis, D. N. & White, H. (2009). Correction to "Automatic
  block-length selection for the dependent bootstrap". *Econometric Reviews*
  28(4), 372-375.
* Politis, D. N. & White, H. (2004). Automatic block-length selection for the
  dependent bootstrap. *Econometric Reviews* 23(1), 53-70.
* Moran, P. A. P. (1950). Notes on continuous stochastic phenomena.
  *Biometrika* 37(1/2), 17-23.
* Nordman, D. J. & Lahiri, S. N. (2004). On optimal spatial subsample size for
  variance estimation. *Annals of Statistics* 32(5), 1981-2027.
* Politis, D. N. & Romano, J. P. (1992). A circular block resampling procedure
  for stationary data. In *Exploring the Limits of Bootstrap*, 263-270. Wiley.
